---
title: "User Guide: 1 Debugging ggplots"
subtitle: "'gginnards' `r packageVersion('gginnards')`"
author: "Pedro J. Aphalo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{User Guide: 3 Debugging ggplots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path = 'figure1/', fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE)
```

## Preliminaries

```{r}
library(ggplot2)
library(gginnards)
library(tibble)
```

We generate some artificial data.

```{r}
set.seed(4321)
# generate artificial data
x <- 1:100
y <- (x + x^2 + x^3) + rnorm(length(x), mean = 0, sd = mean(x^3) / 4)
my.data <- data.frame(x, 
                      y, 
                      group = c("A", "B"), 
                      y2 = y * c(0.5, 2),
                      block = c("a", "a", "b", "b"))
```

We change the default theme to an uncluttered one.

```{r}
old_theme <- theme_set(theme_bw())
```

## Introduction

Package 'ggplot2' defines its own class system, and function `ggplot()` can be
considered as a constructor.

```{r}
class(ggplot())
```

If we pass no arguments an empty plot is constructed.

```{r}
ggplot()
```

The structure of objects of classes `"gg"` `"ggplot"` can be explored with R's 
method `str()` as any other structured R objects. These allows us to see the
different _slots_ of these special type of lists.

```{r}
str(ggplot())
```

If we pass an argument to parameter `data` the data is copied into the
list slot with name `data`. As we also map the data to aesthetics, this
mapping is stored in slot `maaping`. 

```{r}
str(ggplot(data = my.data, aes(x, y, colour = group)))
```

A `summary()` method that produces a more compact output is available in recent
versions of 'ggplot2'. However, it does not reveal the internal structure of
the objects.

```{r}
summary(ggplot(data = my.data, aes(x, y, colour = group)) +
          geom_point())
```

How does mapping work? _Geometries_ (geoms) and _statistics_ (stats) do not "see"
the original variable names, instead the `data` passed to them is named according
to the _aesthetics_ user variables are mapped to. Geoms and stats work in
tandem, with geoms doing the actual plotting and stats summarizing or 
transforming the data. It can be instructive to be able to see what data is
received as input by a geom or stat, and what data is returned by a stat.

Both geoms and stats can have either panel- or group functions. Panel functions
receive as input the subset of the data that corresponds to a whole panel,
mapped to the aesthetics and with factors indicating the grouping (set by
the user by mapping to a discrete scale). Group functions receive as input
the subset of data corresponding to a single group based on the mapping.

The motivation for writing the "debug" stats and geoms included in package
'gginnards' is that at the moment it is in many cases not possible to set
breakpoints inside the code of stats and geoms, because frequently nameless
panel and group functions are stored in list-like objects used to store 
the definitions of geoms and stats.

This can make it tedious to analyse how these functions work, as one may need to
add `print` statements to their source code to see the data. I wrote the "debug"
stats and geoms as tools to help in the development of my packages 'ggpmisc' and
'ggspectra', and as a way of learning myself how data are passed around within
the different components of a `ggplot` object when it is printed.

# Data input to _stats_ 

The code of the _stats_ described in this vignette are very simple and print a summary of
their `data` input by default to the console. However, the default function used to display
the data can be substituted by a different one passed as an argument, adding flexibility. 
The debug stats, in addition return a data frame containing labels suitable for "plotting" debug with geom
"text" or geom "label". 

The 'gginnards' package defines a `"null"` geom, which is used as default 
by the _debug_ _stats_. Currently this geom is similar to the more recently added 
`ggplot2::geom_blank()` and is used as default _geom_ in the _stats_ described in this
user guide.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + 
  geom_null()
```

Using as default geom "null" allows to add the debug _stats_ for the side effect
of console output without altering the the graphic output for the plot when there
is at least one other plot layer.

Because of the way 'ggplot2' works, the values are listed to the console at 
the time when the `ggplot` object is printed. As shown here, no other geom or 
stat is required, however in the remaining examples we add `geom_point()` 
to make the data also visible in the plot.

```{r}
ggplot(my.data, aes(x, y)) + 
  stat_debug_group()
```
 
In the absence of facets or groups we get just get the summary from one data 
frame.

```{r}
ggplot(my.data, aes(x, y)) + 
  geom_point() + 
  stat_debug_group()
```

```{r}
ggplot(my.data, aes(x, y)) + 
  geom_point() + 
  stat_debug_panel()
```

In the case of grouping then one data frame is summarized for each group
in the ggplot object.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + 
  geom_point() + 
  stat_debug_group()
```

Without facets, we still have only one panel.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + 
  geom_point() + 
  stat_debug_panel()
```

The data are similar, except for the column named after the aesthetic, for the aesthetics used for grouping.

```{r}
ggplot(my.data, aes(x, y, shape = group)) + 
  geom_point() + 
  stat_debug_group()
```

# Data returned by _stats_ 

Next we show how `geom_debug()` can be used. Simplest but not most important use is to print to the console the data as passed to geoms as input, but this is not that different from what we saw in the previous section with the debug _stats_.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + 
  geom_point() + 
  geom_debug(summary.fun = head)
```

The main use of `geom_debug()` it to display the data returned by _stats_ and received by the _geoms_. Many but not all extensions to 'ggplot2' document all the computed variables returned by the _stats_ they define through `data`. In addition, when debugging a newly defined _stat_ and/or _geom_ being able to easily see the output is particularly useful when a bug prevents graphical output from being displayed. 

```{r}
ggplot(my.data, aes(x, y, colour = group)) + 
  geom_point() + 
  stat_smooth(method = "lm", geom = "debug")
```

```{r}
ggplot(my.data, aes(group, y, colour = group)) + 
  geom_point(colour = "black") + 
  stat_summary(fun.data = "mean_se") +
  stat_summary(fun.data = "mean_se", geom = "debug")
```

# Grouping and facets

With grouping, for each _group_ the `compute_group()` function is called with a subset of the data.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + 
  geom_point() + 
  stat_debug_group(summary.fun = head, summary.fun.args = list(n = 3))
```

In this example with grouping and facets, within each _panel_ the `compute_group()` function is called for each group, in total four times.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + 
  geom_point() + 
  stat_debug_group(summary.fun = nrow) +
  facet_wrap(~block)
```

With facets, for each _panel_ the `compute_panel()` function is called with a subset of the data that is not split by groups. For our example, it is called twice.

```{r}
ggplot(my.data, aes(x, y, colour = group)) + 
  geom_point() + 
  stat_debug_panel(summary.fun = nrow) +
  facet_wrap(~block)
```

# Controlling the debug output

In the examples above we have demonstrated the use of the _stats_ and _geoms_
using default arguments. Here we show examples of generation of other types of
debug output.

## Display debug output on the plot

If we use as _geom_ `"label"` or `"text"` a _debug_ summary is added
to the plot itself, we can use other arguments valid for the _geom_ used,
in this case `vjust`.

```{r}
ggplot(my.data, aes(x, y, shape = group)) + 
  geom_point() + 
  stat_debug_group(geom = "label", vjust = c(-0.5,1.5))
```

## User summaries for debug output

The default, made explicit.

```{r}
ggplot(my.data, aes(x, y)) + 
  geom_point() + 
  stat_debug_group(summary.fun = as_tibble)
```

If a different summary function is passed as argument to parameter `summary.fun` it will
be used instead of the default one.

```{r}
ggplot(my.data, aes(x, y)) + 
  geom_point() + 
  stat_debug_group(summary.fun = summary)
```

```{r}
ggplot(my.data, aes(x, y)) + 
  geom_point() + 
  stat_debug_group(summary.fun = head)
```

```{r}
ggplot(my.data, aes(x, y)) + 
  geom_point() + 
  stat_debug_group(summary.fun = head, summary.fun.args = list(n = 3))
```

```{r}
ggplot(my.data, aes(x, y)) + 
  geom_point() + 
  stat_debug_group(summary.fun = nrow)
```

This next chunk showing how to print the whole data frame is not run as its output is more than 100 lines long as the data set contains 100 observations.

```{r, eval=FALSE}
ggplot(my.data, aes(x, y)) + 
  geom_point() + 
  stat_debug_group(summary.fun = function(x) {x})
```
## Assign debug output to a variable

In the next example we show how to save the data input of the geom to a variable in the global environment. However, assignment takes place at the time the ggplot object is printed.

```{r}
pipe_assign <- function(value, name, pos = .GlobalEnv, ...) {
  assign(x = name, value = value, inherits = FALSE, pos = pos, ...)
}

ggplot(my.data, aes(x, y, colour = group)) + 
  geom_point() + 
  geom_debug(summary.fun = pipe_assign, 
             summary.fun.args = list(name = "debug_data"),
             print.fun = NULL)

debug_data
```


```{r}
ggplot(my.data, aes(x, y, colour = group)) + 
  geom_point() + 
  geom_debug(summary.fun = NULL)
```



