---
title: "User Guide: 2 Manipulation of plot layers"
subtitle: "'gginnards' `r packageVersion('gginnards')`"
author: "Pedro J. Aphalo"
date: "`r packageDate('gginnards')`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{User Guide: 4 Manipulation of plot layers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path = 'figure2/', fig.align = 'center', 
               fig.show = 'hold', fig.width = 7, fig.height = 4)
options(warnPartialMatchArgs = FALSE)
```

## Introduction

The functions described here are not expected to be useful in everyday plotting
as one can simply change the order in which layers are added to a ggplot.
However, if one uses high level methods or functions that automatically produce
a full plot using 'ggplot2' internally, one may need to add, move or delete
layers so as to profit from such canned methods and retain enough flexibility.

Some time ago I needed to manipulate the layers of a `ggplot`, and found a matching question in [matching question in Stackoverflow](\url{https://stackoverflow.com/questions/13407236/remove-a-layer-from-a-ggplot2-chart}). I used the answers found in Stackoverflow as the starting point for
writing the functions described in this vignette.

In a `ggplot` object, layers reside in a list, and their positions in the list
determine the plotting order when generating the graphical output. The Grammar
of Graphics treats the list of layers as a _stack_ using only _push_ operations.
In other words, always the most recently added layer resides at the end of the
list, and over-plots all layers previously added. The functions described in
this vignette allow overriding the **normal** syntax at the cost of breaking the
expectations of the grammar. These functions are, as told above, to be used only
in exceptional cases. This notwithstanding, they are rather easy to use. The
user interface is consistent across all of them. Moreover, they are designed to
return objects that are identical to objects created using the normal syntax
rules of the Grammar of Graphics.

Function | Use                     
-------- | -------------------------------------
`delete_layers()` | delete one or more layers 
`append_layers()` | append layers at a specific position 
`move_layers()`   | move layers to an absolute position
`shift_layers()`  | move layers to a relative position
`which_layers()`  | obtain the index positions of layers
`extract_layers()`   | extract matched or indexed layers
`num_layers()`    | obtain number of layers
`top_layer()`    | obtain position of top layer
`bottom_layer()`    | obtain position of bottom layer

Although their definitions do not rely on code internal to 'ggplot2', they rely
on the internal structure of objects belonging to class `gg` and `ggplot`.
Consequently, long-term backwards and forward compatibility cannot be
guaranteed, or even expected.

## Preliminaries

```{r}
library(ggplot2)
library(gginnards)
library(tibble)
library(magrittr)
library(stringr)
```

We generate some artificial data.

```{r}
set.seed(4321)
# generate artificial data
my.data <- data.frame(
  group = factor(rep(letters[1:4], each = 30)),
  panel = factor(rep(LETTERS[1:2], each = 60)),
  y = rnorm(40),
  unused = "garbage"
)
```

We change the default theme to an uncluttered one.

```{r}
old_theme <- theme_set(theme_bw())
```

We generate a plot to be used later to demonstrate the use of the functions.

```{r}
p <- ggplot(my.data, aes(group, y)) + 
  geom_point() +
  stat_summary(fun.data = mean_se, colour = "red", size = 1.3) +
  facet_wrap(~panel, scales = "free_x", labeller = label_both)
p
```

## Exploring how ggplots are stored

To display textual information about a `gg` object we use method `summary()`,
while methods `print()` and `plot()` will display the actual plot.

```{r}
summary(p)
```

Layers in a ggplot object are stored in a list as nameless members. This means
that they have to be accessed using numerical indexes, and that we need to use
some indirect way of finding the indexes correspiónding to the layers of
interest.

```{r}
names(p$layers)
```

The output of `summary()` is compact.

```{r}
summary(p$layers)
```

The default `print()` method for a list of layers displays only a small part of 
the information in a layer.

```{r}
print(p$layers)
```

To see all the fields, we need to use `str()`, which we use here for a single layer.

```{r}
str(p$layers[[1]])
```

## Manipulation of plot layers

We start by using `which_layers()` as it produces simply a vector of indexes into
the list of layers. The third statement is useless here, but demonstrates how
layers are selected in all the functions described in this document.

```{r}
which_layers(p, "GeomPoint")
which_layers(p, "StatSummary")
which_layers(p, idx = 1L)
```

We can also easily extract matching layers with `extract_layers()`. Here one layer
is returned, and displayed using the default `print()` method. Method `str()`
can also be used as shown above.

```{r}
extract_layers(p, "GeomPoint")
```

With `delete_layers()` we can remove layers from a plot, selecting them using 
the match to a class, as shown here, or by a positional index. This was shown
above for `which_layers()`.

```{r}
delete_layers(p, "GeomPoint")
```

```{r}
delete_layers(p, idx = 1L)
```

```{r}
delete_layers(p, "StatSummary")
```

With `move_layers()` we can alter the stacking order of layers. The layers to
move are selected in the same way as in the examples above, while `position`
gives where to move the layers to. Two character strings, `"top"` and `"bottom"`
are accepted as `position` argument, as well as `integer`s. In the later case, 
the layer(s) is/are appended after the supplied position with reference to the 
list of layers not being moved. 

```{r}
move_layers(p, "GeomPoint", position = "top")
```

The equivalent operation using a relative position. A positive value for `shift` is 
interpreted as an upward displacement and a negative one as downwards 
displacement. 

```{r}
shift_layers(p, "GeomPoint", shift = +1)
```

Here we show how to add a layer behind all other layers.

```{r}
append_layers(p, geom_line(colour = "orange", size = 1), position = "bottom")

```

It is also possible to append the new layer immediately above an arbitrary
existing layer using a numeric index, which as shown here can be also obtained
by matching to a class name. In this example we insert a new layer in-between
two layers already present in the plot. As with the `+` operator of the Grammar
of Graphics, `object` also accepts a list of layers as argument (no example
shown).

```{r}
append_layers(p, object = geom_line(colour = "orange", size = 1), 
              position = which_layers(p, "GeomPoint"))
```

Annotations add layers, so they can be manipulated in the same way as other layers.

```{r}
p1 <- p + 
  annotate("text", label = "text label", x = 1.1, y = 0, hjust = 0)
p1
```

```{r}
delete_layers(p1, "GeomText")
```

## Other manipulations

We very briefly mention in this section how to manipulate other elements of
of `ggplot` objects. The output from the examples in this section is not
included is this vignette.

For replacing other elements in `ggplot` objects we can in general use methods
defined in 'ggplot2'.

### Replacing scales, coordinates, whole themes and data.

Other elements that are normally _added_ and _replaced_ in plot with operator 
`+`, such as scales. To the default data stored in a `ggplot` object operator 
`%+%` is used. 

### Editing theme elements

Method `summary()` is available for themes.

```{r,eval=FALSE}
summary(theme_bw())
```

However, to see the actual values stored, we need to use `str()`.

```{r, eval=FALSE}
str(theme_bw())
```

Themes can be modified using `theme()`. See the 'ggplot2' documentation for details.

### Removing unused data

When the data frame passed as data contains many columns and rows, it is stored
in whole, even the data columns not mapped to any aesthetic. In many cases this
does not matter, but in the case of big datasets, the use of RAM can rapidly add
up, and in addition the printing of each plot can slow down significantly.

To simplify such objects we need to find which variables are mapped to aesthetics
and which are not. Here is a naive attempt at handling the possibiliy of mappings
to expressions involving computations and multiple variables per mapping, and
facets.

```{r}
mapped.vars <- 
  gsub("[~*\\%^]", " ", as.character(p$mapping)) %>%
  str_split(boundary("word")) %>%
  unlist() %>%
  c(names(p$facet$params$facets))
```

We need also to find which variables are present in the data.

```{r}
data.vars <- names(p$data)
```

Find which variables in data are not used, and delete them from the data.

```{r}
unused.vars <- setdiff(data.vars, c(mapped.vars))
keep.idxs <- which(!data.vars %in% unused.vars)
```

```{r}
p1 <- p
p1$data <- p$data[ , keep.idxs]
```
For a data set this small, removing a single column saves little space.

```{r}
object.size(my.data)
object.size(p)
object.size(p1)
```

The plot has not changed.

```{r}
p1
```

We can assemble everything into a function for convenience. Such a  
function, based on a rough and barely tested piece of code is included in the 
package as `drop_vars()`.

```{r}
drop_vars(p)
```

In principle, one could have expected R not to make a copy of the data in 
memory for each `ggplot` object, however, the ggplot code seems to modify the 
argument passed to `data` triggering a real copy operation. In any case during
printing and when saved to disk, even as part of the environment, avoiding to
carry along unused data can be beneficial. Of course, removing unused data
means that it will not be available if we want to add more layers to the same 
saved ggplot object after dropping unused variables.


